Parametricity Consequences of GADTs Representations

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

@@

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
a b.\,f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $\Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $\Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\mu : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types a calculus must support a
primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
reflection of the higher-order fixpoint operator on the category of
functors on the cagegory interpretating types, as well as a primitive
type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$. {\color{blue} This was elided in ADT
  case. Why? Tie together better.}  The type constructor
$\mathsf{Lan}$ thus computes the smallest data type that is generated
by a GADT's {\color{blue} GADT or constructor?} syntax and also
supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
the functorial completion of the data type determined by a GADT's
syntax.

Note, however, that the functorial completion of a GADT need not be a
nested type. As mentioned in the introduction, only those instances of
the GADT reachable from the syntax of the GADT are in the functorial
completion.

The representation choices are the same for GADTs as for ADTs. The
difference is that, although these two representations are isomorphic
for ADTs, they are not isomorphic for GADTs.  We will see this in
Section~\ref{??} below.

\section{Semantics of GADTs}\label{sec:GADT-semantics}

{\color{blue} OMIT THIS SECTION?}

But the data in a GADT shouldn't disappear just because you map a
function over it!  This is odd! So in what sense to GADTs generalize
ADTs? In what sense are they even data types?

The semantics of GADTs is not so clear.

Functorial completion doesn't give nested types

Functorial completion containerizes the GADT. Extension of a
container. The extension of a container is an endofunctor generalizing
map.

Eq semantics minimal (least commitment, most natural) for CEs; Lan
semantics minimal for primitives. Minimal way to add to any calculus
supporting them.

@@

Since the types encoding GADTs in Bob's system make essential use
of equality types, these considerations extend and possibly generalize
to all GADTs.  Still, the equality is just a shorthand for a type made
by higher-order forall and arrow types, and thus this considerations
are not limited to a specific type, but are intrinsic of Bob's
system.
%We can thus expect that not all types in Bob's system will be
%functorial and that, consequently, the consequences of parametricity
%will not include functoriality but will provide information about,
%e.g., inhabitatedness.  See
%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
%about parametricity and GADTs.

@@

\section{Parametricity}

One important consequence of representation choice is for parametric
reasoning about GADTs. In particular, we can use parametricity to get
our transform-rearrange property for ADTs. Do it for lists?

We can thus expect that not all types in Bob's system will
be functorial and that, consequently, the consequences of
parametricity will not include functoriality but will provide
information about, e.g., inhabitatedness.  See
<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
about parametricity and GADTs.

If we had a parametric model for GADTs, it could perhaps validate
short cut and this show that the two reps for them are equivalent.
(This is a sort of converse.)
The fact that GADTs don't validate any such iso has important
consequences for their parametricity properties.

In this paper we show that there is not an iso between the two reps of
even the simplest GADT {\color{blue} that isn't a hereditarily
  polynomial nested type}, and investigate the consequences of this
discovery for parametric reasoning about GADTs.

Non-uniformity renders build, therefore short cut, for GADTs
impossible. Can't even formulate! Maybe there's another way to prove
equivalence, but I don't know what it would be. In any case, we have a
counterexample, so this settles the question.

@@

Parametric polymorphism as a way to capture Strachey's notion of
uniform, type-independent behavior of functions. Parametric models as
formalizations of this notion of uniformity, as first given in
Reynolds' original paper.

@@

It depends on primitive vs. Church encodings. Primitive forces
functoriality, as in LICS'19.

>>>>

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, a
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. {\color{blue} This is the least commitment
  thing Chirch encodings require.} Such a type can either be taken as
primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

@@

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b. a \times b}\, \lambda
a b. f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $\Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $\Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types a calculus must support a
primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
reflection of the higher-order fixpoint operator on the category of
functors on the cagegory interpretating types, as well as a primitive
type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$. {\color{blue} This is the least commitment
  thing primitive representations can require. Fixpoints were elided
  in ADT case. Why? Tie together better.}  The type constructor
$\mathsf{Lan}$ thus computes the smallest data type that is generated
by a GADT's {\color{blue} GADT or constructor?} syntax and also
supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
the functorial completion of the data type determined by a GADT's
syntax, filling it out with data elements that are derivable via
repeated applications of $\mathsf{map}$ starting from those elements
specified by its syntax. In the primitive view, GADTs are, in general,
underspecified by their syntax.

Just as for ADTs, we can represent GADTs as either Church encodings or
primitive data types. But whereas the two representations are always
isomorphic for ADTs, they not, in general, for GADTs. We will see this
in Section~\ref{??} below. {\color{blue} But we get
  rearrange-transform property for primitive rep since this is a
  consequence of functoriality.}

{\color{blue}
Forced to choose, a programmer would likely find the idea that a GADT
contains data that are not specified by its syntax disturbing. What,
they might ask, should a data type contain other than data that are
constructed using its data constuctors? Why should a GADT contain
``secret'' elements that are only attainable by mapping? On the other
hand, a semanticist would likely find the primitive representation of
GADTs more reasonable. After all, they would likely find the
nonfunctorial nature of the Church encoding unnerving at best. After
all, they would argue, the data in a GADT shouldn't disappear just
because you map a function over it! The fact that this happens for
Church encodings of GADTs --- i.e., for GADTs as specified by their
syntax --- highlights how GADTs {\em do not} generalize the essential
container-ish nature of ADTs at all. A semanticist might therefore
conclude that GADTs are seriously misnamed. They might even go on to
wonder: in what sense are GADTs data types at all?}

\section{Parametricity in the Presence of GADTs}

1. Parametricity requires a fibred semantics. This doesn't happen for
primitive rep.

2. Non-uniformity renders build, therefore short cut, for GADTs
impossible. Can't even formulate! It is possible there could be
another way to prove equivalence, but we know the two representations
are not equivalent because we have a counterexample.

The fact that GADTs don't validate any such iso has important
consequences for their parametricity properties.

@@

Since the types encoding GADTs in Bob's system make essential use
of equality types, these considerations extend and possibly generalize
to all GADTs.  Still, the equality is just a shorthand for a type made
by higher-order forall and arrow types, and thus this considerations
are not limited to a specific type, but are intrinsic of Bob's
system.

@@

We can thus expect that not all types in Bob's system will
be functorial and that, consequently, the consequences of
parametricity will not include functoriality but will provide
information about, e.g., inhabitatedness.
%See
%<https://www.cis.upenn.edu/~sweirich/talks/param-gadt.pdf> and
%<https://www.seas.upenn.edu/~sweirich/papers/rtheorems.pdf> for more
%about parametricity and GADTs.

{\color{blue} Our favorite GADT isn't a hereditarily
  polynomial nested type.}

@@

@@
